\documentclass[11pt]{article}
\usepackage[a4paper,margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,bm}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{hyperref}

\title{Breaking the $\kappa$-collapse: re-anchoring the slow recovery floor to the
DEB maintenance rate constant\\
\large Technical note on the \texttt{lambda\_min} mapping in TwoTimescaleResilience}
\author{Working note}
\date{\today}

\newcommand{\Em}{[E_m]}
\newcommand{\EG}{[E_G]}
\newcommand{\pM}{[p_M]}

\begin{document}
\maketitle

\section{Summary}
The amplification factor $F$ produced by the framework was, under the original
parameterization, numerically a function of the Dynamic Energy Budget (DEB)
\emph{allocation fraction} $\kappa$ \emph{alone}: across the whole AmP library
(7{,}335 species) $\mathrm{Spearman}(F,\kappa)=-1.000$ to machine precision, and the
baseline capacity $A_0$ -- spanning six orders of magnitude -- contributed nothing.
This note shows that the cause is a single mis-normalized parameter: the slow
recovery floor $\lambda_{\min}$ was defined as maintenance power divided by
\emph{reserve density} rather than by the \emph{cost of structure}. Replacing it with
the textbook DEB somatic maintenance rate constant $k_M=\pM/\EG$ both corrects the
quantity and dissolves the collapse: the timescale-separation ratio becomes the
energy investment ratio $g$, and $F$ now tracks $g$ rather than $\kappa$.

\section{The mechanism and the collapse}
The fast (acute) response obeys $\dot y = -\lambda(A)\,y + \beta C_{\mathrm{event}}$,
and a Fubini argument over the event gives the burden $\propto 1/\lambda(A)$, so the
amplification of a chronically-stressed organism relative to a pristine one is
\begin{equation}
F = \frac{\lambda(A_0)}{\lambda(A_t)},
\qquad
\lambda(A) = \lambda_{\min} + (\lambda_{\max}-\lambda_{\min})\,\frac{A_+}{K_A+A_+}.
\label{eq:F}
\end{equation}
$F$ genuinely depends on the current margin $A_t$: this is the model mechanism and is
not in question. The problem is what happens \emph{across species}. The framework
erodes the margin proportionally, $A_t=A_0(1-Q)$ with $Q\in[0,1]$ the fractional
load, and sets $K_A=0.3\,A_0$. Substituting into~\eqref{eq:F}, every factor of $A_0$
cancels:
\begin{equation}
\lambda(A_t)=\lambda_{\min}+(\lambda_{\max}-\lambda_{\min})\frac{1-Q}{1.3-Q},
\qquad
\lambda(A_0)=\lambda_{\min}+\frac{\lambda_{\max}-\lambda_{\min}}{1.3},
\end{equation}
so, with $r\equiv\lambda_{\max}/\lambda_{\min}$,
\begin{equation}
F(Q)=\frac{1+(r-1)/1.3}{\,1+(r-1)\dfrac{1-Q}{1.3-Q}\,}.
\label{eq:FQ}
\end{equation}
$F$ is a function of $(r,Q)$ only. The capacity scale $A_0$ has vanished. Everything
therefore hinges on the ratio $r$.

\subsection{Why $r=1/\kappa$ in the original mapping}
The offline mapping (\texttt{src/AmP\_Translator.jl}) set
\begin{equation}
\lambda_{\max}=\frac{v}{L_m},
\qquad
\lambda_{\min}=\frac{\pM}{A_0}=\frac{\pM}{\Em},
\end{equation}
with $A_0=\Em=p_{Am}/v$ and $L_m=\kappa\,p_{Am}/\pM$. Then
\begin{equation}
r=\frac{\lambda_{\max}}{\lambda_{\min}}
 =\frac{v/L_m}{\pM/\Em}
 =\frac{v\,\pM/(\kappa\,p_{Am})}{\pM\,v/p_{Am}}
 =\frac{1}{\kappa}.
\end{equation}
The cancellation of $p_{Am}$, $\pM$ and $v$ is complete; only the allocation fraction
survives. Combined with~\eqref{eq:FQ} this makes $F$ a one-parameter family in
$\kappa$. The capacity signal the framework advertises ($A_0$, the ``evolutionary
buffer'') is inert.

\section{Diagnosis: $\lambda_{\min}$ is not the maintenance rate constant}
$\lambda_{\min}=\pM/\Em$ is maintenance power normalized by \emph{reserve density}.
That is not a standard DEB quantity. The canonical \emph{somatic maintenance rate
constant} is maintenance power normalized by the \emph{volume-specific cost of
structure},
\begin{equation}
k_M=\frac{\pM}{\EG},
\end{equation}
which is the fundamental slow rate constant of DEB theory (it sets the von Bertalanffy
and maturation timescales). Normalizing by $\Em$ rather than $\EG$ drags $p_{Am}$ and
$v$ into $\lambda_{\min}$ in precisely the way that forces $r=1/\kappa$.

\section{The fix}
Keep $\lambda_{\max}=v/L_m$ and set the floor to the maintenance rate constant, with a
clamp to preserve $\lambda_{\min}\le\lambda_{\max}$:
\begin{equation}
\boxed{\;\lambda_{\min}=\min\!\bigl(k_M,\;\lambda_{\max}\bigr),\qquad k_M=\frac{\pM}{\EG}.\;}
\end{equation}
Using the standard DEB identity $L_m=v/(g\,k_M)$, the timescale-separation ratio for
unclamped species becomes the \emph{energy investment ratio}
\begin{equation}
r=\frac{\lambda_{\max}}{\lambda_{\min}}=\frac{v/L_m}{k_M}=g=\frac{\EG}{\kappa\,\Em},
\end{equation}
a primary DEB parameter that varies across species by the reserve/structure economy,
not merely by allocation. With $\lambda_{\min}=k_M$ and $\lambda_{\max}=g\,k_M$ the two
rates share the metabolic tempo $k_M$, which \emph{cancels} in~\eqref{eq:F}: $k_M$ sets
how \emph{fast} recovery is, $g$ sets how \emph{much} amplification there is.

The clamp encodes a deliberate modelling choice: reserve-rich species ($g<1$, hence
$k_M>v/L_m$) receive no timescale separation, so $\lambda$ is constant and $F\equiv1$.
Such species are resilient to chronic-then-acute stress by construction.

Both $k_M$, $\EG$ and $g$ are read directly from AmP (\texttt{allStat.mat}); the change
is otherwise local to \texttt{AmP\_Translator.jl}, and the species library JSON is
regenerated from it.

\section{Empirical result}
Re-running over the regenerated library (7{,}335 species;
\texttt{examples/amp\_lambda\_min\_validation.jl}):

\begin{center}
\begin{tabular}{lcc}
\toprule
quantity & before fix & after fix \\
\midrule
$\mathrm{Spearman}(F,\kappa)$ & $-1.000$ & $-0.107$ \\
$\mathrm{Spearman}(F,g)$      & ---      & $+0.725$ \\
$\mathrm{Spearman}(g,\kappa)$ & ---      & $-0.244$ \\
$\max|\,\lambda_{\max}/\lambda_{\min}-g\,|$ (unclamped) & --- & $2\times10^{-13}$ \\
\bottomrule
\end{tabular}
\end{center}

The collapse is broken: $F$ is no longer a $\kappa$ ranking, it tracks $g$, and $g$ is
only weakly correlated with $\kappa$ ($-0.244$) so the change genuinely adds
physiological information. The identity $r=g$ holds to machine precision.

\section{Newly exposed question: the range of $g$}
The fix is directionally correct but reveals the next design problem. Because $g$ in
AmP spans roughly $10^{-3}$ to $\sim\!10^{2}$:
\begin{itemize}[nosep]
  \item $65\%$ of species have $g\le1$ and are clamped to $F\equiv1$ (binary resilience);
  \item at full margin collapse ($Q\to1$) the ceiling $F_{\max}=1+(g-1)/1.3$ reaches
        $\sim\!559$ for extreme-$g$ species.
\end{itemize}
At moderate erosion the picture is tame ($F\le1.23$ at $Q=0.5$): the explosion is only
the $A_t\to0$ limit. Still, mapping the recovery-rate ratio to $g$ \emph{raw} is
probably too literal. Open options, to be weighed with DEB domain input:
\begin{enumerate}[nosep]
  \item accept the clamp ($g<1\Rightarrow$ resilient) and bound the upper tail with a
        saturating transform of $g$;
  \item reconsider $\lambda_{\max}=v/L_m$ (the slowest reserve turnover, evaluated at
        $L_m$) as the fast reference rate;
  \item treat $A_0$ as an absolute (not relative) margin scale by breaking
        $K_A\propto A_0$ -- a separate and, per the structure diagnostic, weaker lever.
\end{enumerate}

\section{Status}
This change is committed on the branch
\texttt{fix/lambda-min-maintenance-rate}. It is a principled correction (the previous
$\lambda_{\min}$ was the wrong rate constant) rather than a tuning choice, and the
$\kappa$-collapse dissolving is the signal that the original normalization was wrong.
Whether $g$-driven vulnerability (reserve-poor, structure-expensive species amplify
more) is biologically correct, and how to tame the range of $g$, are the questions for
empirical validation and DEB expert review. Companion material:
\texttt{docs/wiki/Limitations-and-Open-Questions.md},
\texttt{examples/amp\_kappa\_collapse\_diagnostic.jl},
\texttt{examples/amp\_lambda\_structure\_comparison.jl}.

\end{document}
